Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-08-19 review_article.Rmd
Add figure reference

The visualisation methods used to present cancer statistics will depend on the intended message and users of the map. 
Presenting statistics as visualisations requires transforming individual observations into aggregations of communities as geographical units.
Visualising diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. 
Disease maps help to present geographic patterns that may have been overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease are able to be identified and effectively addressed.

@@

<!-- what is a choropleth map -->
A choropleth map is used to display the characteristics of a spatial relationship of measurements. 
A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns on a familiar geographic form. 
Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organising communities into representable units [@SAMGIS]. 
A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI].
Early versions of choropleth maps used symbols or patterns instead of colour.
As an alternative storage device to a table, it preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO].
@CPISACA discuss the use of choropleths to visualise cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.

@@

@@

Examples of atlases:
Kraak 1998, Kraak and Ormeling 2011
Bertin 1967

Current atlases utilising choropleth methods 
- include Australia:

cites: Cancer Mortality Atlas of EU and of the European Economic Area last published 8 years ago
@MTMSIH Bayesian disease mapping methods for areal data - explains key concepts applied to create an atlas
Explores Australia as a case study, utilising QLD SA2 areas

@@

To be able to recognise the significant changes, a reader will usually have to know the initial geography to find the differences in the new cartogram layout [@NAC]. Tobler's Conformal mapping means to preserve angles
locally so that the shapes of very small areas on a traditional map and a cartogram would be similar.
@CBATCC presents this issue as conflicting tasks or aims, to adjust region sizes and retain region shapes. Distortion of region shapes on the contiguous cartogram presents an additional hurdle to visual recognition and this hurdle is not only eliminated on the noncontiguous cartogram but is replaced by the meaningful empty-space property [@NAC, @ECGC].

Tobler's method and the many implementations that 'elaborated' on it are derived from 'numerical approximations to a pair of equations'[@ACTUC]. They all operate through incremental adjustments, and can produce wildly different outcomes from small changes in the inputs.

@TFYCC Value-Area Cartograms. In these cartograms a region,country, or continent is subdivided into small regions, each of which is represented by a rectangle. This rectangle is proportionate in area to the value which it represents in certain statistical distributions. The regions are grouped in approximately the same positions as they are on the map.

@@

The shapes are valuable for recognition and allows users to orient themselves on the display. Map creators can efficiently communicate with this kind of map by keeping the outlines or particular elements of the original in the new shape [@NISCC].
The scale of the areas does not impact on the shape recognition. However it may impact on the visibility of all areas if small areas expand beyond their boundaries.

@@

### Dorling

@@

### Alternative displays

Original value by area cartograms, especially the tile method, laid the foundation for the modern alternative displays.

@MDAC define the term 'mosaic cartograms'. 
Compare amount of tiles to contrast population of regions.
'Cartograms show a data value per input region by scaling each region such that its area is proportional to its data value. Mosaic cartograms show data in multiples of tiles, hence the input data must consist of, or be cast into, small integer units.'

@@

<!--
@SE Researchers mapping spatial patterns of disease
for small areas are confronted with a number of
challenges, including cartographic design, the
modifiable area unit problem, and also the “small
number” problem, where some places have no,
or few, disease events and/or a low population at
risk of contracting the disease.

>>>>

The visualisation methods used to present cancer statistics will depend on the intended message and users of the map. 
Presenting statistics as visualisations requires transforming individual observations into aggregations of communities as geographical units.
Visualising diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. 
Disease maps help to present geographic patterns that may have been overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease are able to be identified and effectively addressed.
@SE recognises one of the key challenges with mapping spatial patterns of disease is the design of visualisations.
This paper addresses the visualisation techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

@@

<!-- what is a choropleth map -->
A choropleth map is used to display the characteristics of a spatial relationship of measurements. 
A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns on a familiar geographic form. 
Figure \ref{fig:choro} shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organising communities into representable units [@SAMGIS]. 
A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI].
Early versions of choropleth maps used symbols or patterns instead of colour.
As an alternative storage device to a table, it preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO].
@CPISACA discuss the use of choropleths to visualise cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.

@@

The atlas of Cancer in Queensland [@DACQ] focussed on highlighting the difference in experience for those living in rural and disadvantaged areas. They used Standardised Incidence Ratios modelled via Bayesian methods, and explored the information presented in a range of selected atlases. These atlases provided Incidence, Survival and Mortality rates to the public on Areal or Isopleth map bases.

The presentation of these rates considered not only the map base, but also appropriate choices of colour blind friendly colour schemes, and categories of the values to highlight signifcantly different areas.

@@

Examples of atlases:
Kraak 1998, Kraak and Ormeling 2011
Bertin 1967
Coleman MP, Gatta G, Verdecchia A, Esteve J, Sant M, Storm H, Allemani C,Ciccolallo L, Santaquilani M, Berrino F:EUROCARE-3 summary: cancersurvival in Europe at the end of the 20th century.Ann Oncol2003,14(Suppl 5):v128-149

National Cancer Institute:Cancer Trends Progress Report - 2009/2010 UpdateBethesda: NCI, NIH, DHHS; 2010

Pickle LW:A history and critique of U.S. mortality atlases.Spatial andSpatio-temporal Epidemiology2009,1:3-17.

Mason TJ, McKay FW, Hoover R, Blot WJ, Fraumeni JF:Atlas of cancermortality for U.S. counties: 1950-1969Washington: U.S. Govt. Printing Office;1975.
howmick T, Robinson AC, Gruver A, MacEachren AM, Lengerich EJ:Distributed usability evaluation of the Pennsylvania Cancer Atlas.Int JHealth Geogr2008,7:36.

Queensland Cancer Registry:Cancer in Queensland: Incidence, Mortality,Survival and Prevalence,1982 to 2007Brisbane: QCR, Cancer CouncilQueensland and Queensland Health; 2010.

New York State Department of Health Maps of Cancer Incidence byCounty.[http://www.health.state.ny.us/statistics/cancer/registry/cntymaps/index.htm].
Quinn M, Wood H, Cooper N, Rowan S, Eds:Cancer Atlas of the UnitedKingdom and Ireland 1991-2000.London: Office for National Statistics;2005.
-->

@@

To be able to recognise the significant changes, a reader will usually have to know the initial geography to find the differences in the new cartogram layout [@NAC]. Tobler's Conformal mapping means to preserve angles
locally so that the shapes of small areas on a traditional map and a cartogram would be similar.
@CBATCC presents this issue as conflicting tasks or aims, to adjust region sizes and retain region shapes. 
@MDAC define the term 'mosaic cartograms' for hexagonal tile displays, where the amount of tiles for each area can be used to communicate the statistic of regions. 
The complexity of the boundaries can be adjusted in the resulting display, as the size of the tiles used allows a trade-off be made between boundary complexity and simplicity.

<!--
Tobler's method and the many implementations that 'elaborated' on it are derived from 'numerical approximations to a pair of equations'[@ACTUC]. They all operate through incremental adjustments, and can produce wildly different outcomes from small changes in the inputs.

@@

As the trade-off regarding boundaries approaches simplicity, the distortion of region shapes on the contiguous cartogram presents an additional hurdle to visual recognition and this hurdle is not only eliminated on the noncontiguous cartogram but is replaced by the meaningful empty-space property [@NAC, @ECGC].
The shapes are valuable for recognition and allows users to orient themselves on the display. Map creators can efficiently communicate with this kind of map by keeping the outlines or particular elements of the original in the new shape [@NISCC].
The scale of the areas does not impact on the shape recognition. However it may impact on the visibility of all areas if small areas expand beyond their boundaries.

@@

### Alternative Displays

@@

The groundwork for the simplistic approach had been laid in the mid 1930's be Raisz. @TFYCC quotes the official definition of Value-Area Cartograms, the simplistic displays which represent each area as a single rectangle, sized according to the value of the statistic.
This rectangluar display also allows for tiling, where geographic neighbours also share rectangle borders.

@@

<!--
-->
